Approximate Lattice Matching in Three Dimensions

### 2025-11-23 3D_Niggli_Matrices_uv5039/3D_Niggli_Matrices_uv5039.tex
just saving work

\keyword{lattice}
\keyword{reduction}
\keyword{Niggli}

	\maketitle
	\begin{abstract}
		Niggli reduction finds the most compact unit cell of a 3-D lattice.  
		Existing algorithms operate in \GVI{} space, and the 3-D basis-vector changes are lost.  
		We tabulate the 3×3 matrices for every step of the reduction, enabling direct re-indexing of reflections and full reconstruction of the reduction history.

@@

		\begin{synopsis}

	\end{synopsis}

@@

\section{Synopsis}

Niggli reduction is formulated in terms of 6 scalars that formally are the components of space \GVI{}. Working in a
higher dimensional space loses the 3-space transformations that
are needed for reindexing reflection data and for changing
the orientation of atomic structure. The matrices are listed
here corresponding to each of the operations of Niggli reduction.

>>>>

\keyword{lattice}
\keyword{reduction}
\keyword{Niggli}
\keyword{matrices}
	\maketitle
	\begin{abstract}
		Niggli reduction finds the most compact unit cell of a 3-D lattice.  
		Existing algorithms operate in \GVI{} space, and the 3-D basis-vector changes are lost.  
		We tabulate the 3×3 matrices for every step of the reduction, enabling direct re-indexing of reflections and full reconstruction of the reduction history.

@@

@@

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.

\section{Availability of code} Code for TransformationMatrices.h is available in github.com, in
\url{https://github.com/duck10/LatticeRepLib.git}.

%\section{Synopsis}
%
%Niggli reduction is formulated in terms of 6 scalars that formally are the components of space \GVI{}. Working in a
%higher dimensional space loses the 3-space transformations that
%are needed for reindexing reflection data and for changing
%the orientation of atomic structure. The matrices are listed
%here corresponding to each of the operations of Niggli reduction.

### 2025-11-23 LatticeMatching_2025/LatticeMatching_2025.tex
just saving work

@@

		{\bf Note:}  In his later publications, Boris Delaunay used the Russian version of his surname, Delone.\\

@@

	A relatively common crystallographic problem is to compare two 
	unit cells and determine whether their lattices are the or
	close to each other. \citeasnoun{flor2016comparison} have described a
	solution for more symmetric crystal families. However, for the general case of triclinic lattices
	a different method is required.

@@

	 Attempts to solve the problem
	by trying a large number of trial transformations can require
	an unknown and unknowable number of trials of transformation. Here we describe
	a straight-forward approach that uses a fixed, modest number
	of trials and that gives multiple possible matches if they are
	appropriate.

@@

	The general solution to matching two lattice has been to 
	test the comparison using a large number of likely transformations.
	Two problems arise: how many trials to make and how to effectively
	measure the difference between two lattices.

	To avoid the need to examine an unknown number of possible 
	transformations, we propose an algorithm with the following stages.
	One of the input cells will be designated ``reference'' and the
	other (to which the transformations in stage three will be applied) will be
	designated ``mobile''.
	\begin{enumerate}
		\item Convert from both (possibly) centered lattices to primitive.
		(Apply matrix $M_{CR}$ for reference and $M_{CM}$ for mobile.) \cite{ITC_VolumeA_2016_A}
		\item Niggli reduce both lattices. (Apply matrix $M_{NR}$ for reference and $M_{NM}$ for mobile.) 
		\item Apply a limited set of transformations to mobile, measuring an
		agreement factor for each trial, keeping those that meet
		an acceptable agreement value. This stage creates matrix $M_{MR}$
		that transforms the modified mobile to the modified reference. \cite{Niggli1928} \cite{Andrews2025b} 
		\item Compare the results of each transformation using
		the \PIII{} Euclidean distance between the reference and the mobile 
		cell. \cite{Andrews2025c} 
	\end{enumerate}

@@

	The immediate question is ``how many transformations will
	need to be tested in stage three''. We have chosen to use all
	the unimodular matrices ({\it i.e.} determinant = 1.0), with
	elements -1, 0, +1. (3480 total matrices). Each transformation
	is tested and acceptable results are accumulated to be ``finished''.

@@

	\subsection{example 1}
{
	\noindent personal communication, \citeasnoun{aroyo2025pc}\\
	More than one acceptable match was found\\
\noindent Test input\\
p 5.17 3.18 7.74 90. 104.5 90.  (reference)\\
p 30.9616 3.1800 8.1608 90.00 171.18 90.00 (mobile)\\

@@

\subsection{example 3}
	\noindent personal communication 
	\cite{simmons2025pc}\\
	A case where more than one accepted match was found.\\
	Input:\\
	p 10.25 10.74 21.08 87.72 75.97 61.53 (reference)\\
	p 10.25 10.74 21.08 78.96 75.97 61.49 (mobile)\\

\begin{verbatim}
		--- Match 1 ---
	Quality: EXCELLENT (0.010 )
	P3 Distance: 0.010
	S6 Angle: 0.01
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[ -1.0000,   1.0000,   0.0000]
	[  0.0000,   0.0000,  -1.0000]
	Matrix determinant: 1.0000
	    Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
			Transformed cell: P    10.250    10.739    21.080    87.714    75.970    61.502
	*** SUCCESS: Excellent match within P3-relative threshold ***

	--- Match 2 ---
	Quality: EXCELLENT (0.023 )
	P3 Distance: 0.023
	S6 Angle: 0.03
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[  0.0000,  -1.0000,   0.0000]
	[ -1.0000,   0.0000,   1.0000]
Matrix determinant: 1.0000
    	Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)\\
    	Transformed cell: P    10.250    10.740    21.087    87.674    75.893    61.490
*** SUCCESS: Excellent match within P3-relative threshold ***

@@

\subsection{example 4}{
%%	C:\Users\lca\Source\Repos\duck10\LatticeRepLib\x64\Release>cmdgen 1 C5 | cmdperturb 20 | cmdniggli | cmdtocell
%	; To Cell
%	; Enter control variables and input vectors (type 'end' to finish):
%	end
20 slight variations on a primitive cubic cell
were
created, with a=10.0, each Niggli reduced, and lattice-matched.\\

Input:
\begin{verbatim}
	P   7.579   7.579   7.579  90.000  90.000  90.000 (reference)
	P   7.570   7.573   7.574  89.957  89.937  89.958
	P   7.576   7.580   7.581  89.989  89.957  89.964
	P   7.573   7.574   7.576  89.950  89.946  89.994
	P   7.571   7.578   7.583  90.058  90.029  90.062
	P   7.577   7.578   7.578  90.051  90.033  90.051
	P   7.572   7.576   7.578  90.033  90.034  90.084
	P   7.573   7.576   7.580  90.028  90.060  90.020
	P   7.570   7.573   7.580  89.993  89.977  89.931
	P   7.572   7.572   7.573  89.982  89.934  89.936
	P   7.577   7.580   7.588  89.958  89.970  89.996
	P   7.571   7.581   7.581  90.061  90.028  90.059
	P   7.573   7.574   7.584  90.005  90.023  90.065
	P   7.580   7.581   7.581  89.933  89.996  89.968
	P   7.576   7.581   7.586  89.913  89.990  89.956
	P   7.573   7.574   7.581  89.979  89.983  89.974
	P   7.570   7.575   7.57   90.025  90.063  90.066
	P   7.571   7.579   7.581  89.988  89.993  89.942
	P   7.578   7.581   7.588  90.049  90.023  90.002
	P   7.580   7.580   7.585  89.932  89.989  89.988
	P   7.571   7.576   7.577  89.970  89.993  89.943
\end{verbatim}

\begin{verbatim}
		and here is the matching results:

	Most common transformation matrices:
	[1,0,0,0,1,0,0,0,1]   : 9 mobiles (P3 agreement: 0.016)
	[1,0,0,0,-1,0,0,0,-1] : 3 mobiles (P3 agreement: 0.015)
	[0,1,0,0,0,-1,-1,0,0] : 2 mobiles (P3 agreement: 0.015)
	[1,0,0,0,0,-1,0,1,0]  : 1 mobiles (P3 agreement: 0.011)
	[0,0,1,1,0,0,0,1,0]   : 1 mobiles (P3 agreement:, 0.012)
	[0,0,1,0,-1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.014)
	[0,0,-1,0,1,0,1,0,0]  : 1 mobiles (P3 agreement: 0.012)
	[0,-1,0,-1,0,0,0,0,-1]: 1 mobiles (P3 agreement: 0.009)
	[-1,0,0,0,0,-1,0,-1,0]: 1 mobiles (P3 agreement: 0.017)

	=== BEST MATCH DETAILS ===
	Mobile 2: 8.42e-03 (EXCELLENT)
	Matrix: [1 0 0 0 1 0 0 0 1]
	Reference	 P     7.579     7.579     7.579    90.000    90.000    90.000
	Transformed:     7.576     7.580     7.581    89.989    89.957    89.964

	Mobile 15: 9.36e-03 (EXCELLENT)
	Matrix: [0 -1 0 -1 0 0 0 0 -1]
	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
	Transformed:     7.574     7.573     7.581    89.983    89.979    89.974

	Mobile 13: 1.03e-02 (EXCELLENT)
	Matrix: [1 0 0 0 1 0 0 0 1]
	Reference:   P   7.579     7.579     7.579    90.000    90.000    90.000
	Transformed      7.580     7.581     7.581    89.933    89.996    89.968

@@

@@

	The code for Lattice Matching and related 
	software tools are available in github.com, in
	\url{https://github.com/duck10/LatticeRepLib.git}.
	The program CmdLMP3 uses the required files.

@@

	%\section{blah blah blah -- Supplementary Material}
	\ack{Acknowledgements}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged.

%	Our thanks to Jean Jakoncic and Alexei Soares for 
%	helpful conversations and access to data and facilities at 
%	Brookhaven National Laboratory.
%	
\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and 
Environmental Research and of Basic Energy Sciences 
(grant No. DE-AC02-98CH10886; grant No. E-SC0012704); 
U.S. National Institutes of Health (grant No. P41RR012408; 
grant No. P41GM103473; grant No. P41GM111244; 
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd.

@@

A solution is proposed for the problem of best matching the 
lattices of unit cells that are described in different
presentations.

>>>>

%Andrea Hill
%2:18 AM (4 hours ago)
%to Herbert, me, Mois
%
%Dear Larry and Herbert
%
%Thanks for the submissions uv5038 and uv5039.
%
%Please cite these in the format Authors, Year, Journal, Volume, Note, ie:
%
%Andrews, L. & Bernstein, H. J. (2025a). Acta Cryst. A81. Submitted (uv5038)
%
%Andrews, L. & Bernstein, H. J. (2025b). Acta Cryst. A81. Submitted (uv5039)
%
%You could also give the article code (uv5038/uv5039) in plain text in the citations, and provide the article title for reviewers.
%
%Best wishes, Andrea

@@

@@

	A relatively common crystallographic problem is to compare two 
	unit cells and determine whether their lattices are the same or
	close to each other. \citeasnoun{flor2016comparison} have described a
	solution for more symmetric crystal families. However, for the general case of triclinic lattices
	a different method is required.

@@

	\citeasnoun{Mighell1996} give a flow chart that partly encompasses the
	method here described, but no details are given there. Whereas their
	method relies on completely on Niggli reduction, our method does
	not, only including Niggli reduction as part of the processes.
	In addition, their description of the generation of alternative
	unit cells is quite brief, focusing on subcell and supercell
	transformations.

	 Attempts to solve the problem
	by trying a large number of trial transformations can require
	an unknown and unknowable number of trials of transformations. Here we describe
	a straight-forward approach that uses a fixed, modest number
	of trials and that gives multiple possible matches if they are
	near or identical to the best match.

@@

	The general solution approach to matching two lattices has been to 
	test the comparison using a large number of likely transformations.
	Two problems arise: how many trials to make and how to effectively
	measure the difference between two lattices.

	To avoid the need to examine an unknown number of possible 
	transformations, we propose an algorithm with the following stages.
	One of the input cells will be designated ``reference'' and the
	other (to which the transformations in stage three will be applied) will be
	designated ``mobile''.
	\begin{enumerate}
		\item Convert from both (possibly) centered lattices to primitive.
		(Apply matrix $M_{CR}$ for reference and $M_{CM}$ for mobile.) \cite{ITC_VolumeA_2016_A}
		\item Niggli reduce both lattices. (Apply matrix $M_{NR}$ for reference and $M_{NM}$ for mobile.) 
		\item Apply a limited set of transformations to mobile, measuring an
		agreement factor for each trial, keeping those that meet
		an acceptable agreement value. This stage creates matrix $M_{MR}$
		that transforms the modified mobile to the modified reference. \cite{Niggli1928} \cite{Andrews2025b} 
		\item Compare the results of each transformation using
		the \PIII{} Euclidean distance between the reference and the mobile 
		cell. \cite{Andrews2025a} 
	\end{enumerate}

@@

	The immediate question is ``how many transformations will
	need to be tested in stage three''. We have chosen to use all
	the unimodular matrices ({\it i.e.} determinant = 1.0), with
	elements -1, 0, +1. (3480 total matrices, by direct computation). Each transformation
	is tested and acceptable results are accumulated to be ``finished''.

@@

	\subsection{example 1}
{

	More than one acceptable match was found\\
\noindent Test input\\
p 5.17 3.18 7.74 90. 104.5 90.  (reference)\\
p 30.9616 3.1800 8.1608 90.00 171.18 90.00 (mobile)\\

@@

\subsection{example 3}

	A case where more than one accepted match was found. ()Personal communication: \citeasnoun{simmons2025pc}.)\\
	Input:\\
	p 10.25 10.74 21.08 87.72 75.97 61.53 (reference)\\
	p 10.25 10.74 21.08 78.96 75.97 61.49 (mobile)\\

\begin{verbatim}
		--- Match 1 ---
	Quality: EXCELLENT (0.010 )
	P3 Distance: 0.010
	S6 Angle: 0.01
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[ -1.0000,   1.0000,   0.0000]
	[  0.0000,   0.0000,  -1.0000]
	Matrix determinant: 1.0000
	    Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)
			Transformed cell: P    10.250    10.739    21.080    87.714    75.970    61.502
	*** SUCCESS: Excellent match within P3-relative threshold ***

	--- Match 2 ---
	Quality: EXCELLENT (0.023 )
	P3 Distance: 0.023
	S6 Angle: 0.03
	Transformation Matrix:
	[ -1.0000,   0.0000,   0.0000]
	[  0.0000,  -1.0000,   0.0000]
	[ -1.0000,   0.0000,   1.0000]
Matrix determinant: 1.0000
    	Reference cell    p    10.25     10.74     21.08     87.72     75.97     61.53 (reference)
    	Transformed cell: P    10.250    10.740    21.087    87.674    75.893    61.490
*** SUCCESS: Excellent match within P3-relative threshold ***

@@

\subsection{example 4}{
%%	C:\Users\lca\Source\Repos\duck10\LatticeRepLib\x64\Release>cmdgen 1 C5 | cmdperturb 20 | cmdniggli | cmdtocell
%	; To Cell
%	; Enter control variables and input vectors (type 'end' to finish):
%	end
Twenty slight variations of a primitive cubic unit cell (a = 10.0), expressed in \SVI{}, were generated by adding random orthogonal perturbation vectors to the original \SVI{} representation. Each perturbation had a length of 0.001 times the \SVI{} norm of the cubic cell and was added to the vector [0, 0, 0, –100, –100, –100]. The resulting \SVI{} vectors were then converted back to unit cell parameters.
 The modified vectors
were lattice-matched to the reference, which is the first entry.\\

Input:
\begin{verbatim}
P    10.000    10.000    10.000    90.000    90.000    90.000 Reference
P     9.994     9.992     9.988    89.958    89.937    89.957
P    10.002    10.003     9.996    89.957    90.036    90.011
P     9.996     9.992     9.993    89.994    89.950    89.946
P     9.990     9.999    10.005    90.058    89.971    89.938
P     9.997     9.999     9.998    90.051    89.949    89.967
P     9.996     9.998     9.990    89.966    89.916    90.033
P    10.002     9.993     9.996    89.980    90.028    89.940
P    10.002     9.992     9.988    89.931    89.977    89.993
P     9.991     9.992     9.991    89.934    89.936    89.982
P    10.002     9.998    10.012    90.030    90.042    89.996
P    10.003     9.990    10.002    89.972    90.061    89.941
P     9.993    10.006     9.992    90.023    89.935    89.995
P    10.002    10.002    10.002    89.968    90.004    90.067
P    10.003    10.010     9.996    90.010    89.956    90.087
P     9.992    10.003     9.994    89.979    89.974    89.983
P     9.999     9.995     9.988    89.934    89.937    90.025
P    10.003     9.990    10.000    89.942    90.012    90.007
P    10.012     9.998    10.002    90.002    90.049    90.023
P    10.008    10.002    10.002    90.012    90.068    89.989
P     9.989     9.996     9.998    89.970    89.993    89.943
\end{verbatim}

\begin{verbatim}
		and here are the matching results:

Most common transformation matrices:
[-1,0,0,0,-1,0,0,0,1] : 3 mobiles (EXCELLENT, 0.019)
[1,0,0,0,-1,0,0,0,-1] : 2 mobiles (EXCELLENT, 0.013)
[0,0,-1,0,-1,0,-1,0,0]: 2 mobiles (EXCELLENT, 0.022)
[0,-1,0,-1,0,0,0,0,-1]: 2 mobiles (EXCELLENT, 0.020)
[-1,0,0,0,1,0,0,0,-1] : 2 mobiles (EXCELLENT, 0.014)
[-1,0,0,0,0,-1,0,-1,0]: 2 mobiles (EXCELLENT, 0.022)
[1,0,0,0,1,0,0,0,1]   : 1 mobiles (EXCELLENT, 0.016)
[0,1,0,1,0,0,0,0,-1]  : 1 mobiles (EXCELLENT, 0.015)
[0,1,0,0,0,1,1,0,0]   : 1 mobiles (EXCELLENT, 0.017)
[0,1,0,-1,0,0,0,0,1]  : 1 mobiles (EXCELLENT, 0.016)

=== BEST MATCH DETAILS ===
Mobile 2: 1.13e-02 (EXCELLENT)
Matrix: [0 0 -1 1 0 0 0 -1 0]
Transformed:     9.996    10.002    10.003    89.989    89.957    89.964

Mobile 15: 1.23e-02 (EXCELLENT)
Matrix: [-1 0 0 0 0 -1 0 -1 0]
Transformed:     9.992     9.994    10.003    89.979    89.983    89.974

Mobile 13: 1.34e-02 (EXCELLENT)
Matrix: [1 0 0 0 -1 0 0 0 -1]
Transformed:    10.002    10.002    10.002    89.968    89.996    89.933

\subsection{Example 5}

Two reported polymorphic forms of triphenylphosphine oxide
hemihydrate, C\textsubscript{18}H\textsubscript{15} PO$\cdot$0.5H\textsubscript{2}O. \cite{mighell2011role}

\begin{verbatim}
	F 19.794 32.54 9.459 90 90  90
	C 9.4313 32.193 10.8435 90 115.742 90
	end

	=== PROCESSING INPUT LIST ===
	Total results before processing: 500
	NC Distance: 6.44756
	Range: 4.34e-01 to 6.61e+01 
	Reference P3 norm: 39.244 
	Method: Fixed strict
	Thresholds:
	EXCELLENT <= 1.96e-01 (0.5% of P3)
	GOOD      <= 7.85e-01 (2.0% of P3)
	POOR      <= 3.14e+00 (8.0% of P3)

	=== LATTICE MATCHING RESULTS === ===================================================
	Found 2 good matches
	Success threshold: 1.96e-01  (Fixed strict)

	--- Match 1 ---
	Quality: GOOD (0.434 )
	P3 Distance: 0.434
	S6 Angle: 0.12 degrees
	Transformation Matrix:
	[ -1.0000,   0.0000,  -2.0000]
	[  0.0000,  -1.0000,   0.0000]
	[ -1.0000,   0.0000,   0.0000]
	Matrix determinant: 2.0000
	Transformed cell: F    19.535    32.193     9.431    90.000    89.964    90.000
	*** SUCCESS: Good match within P3-relative threshold ***

	--- Match 2 ---
	Quality: GOOD (0.434 )
	P3 Distance: 0.434
	S6 Angle: 0.12 degrees
	Transformation Matrix:
	[  1.0000,   0.0000,   2.0000]
	[  0.0000,  -1.0000,   0.0000]
	[  1.0000,   0.0000,   0.0000]
	Matrix determinant: 2.0000
	Transformed cell: F    19.535    32.193     9.431    90.000    89.964    90.000
	*** SUCCESS: Good match within P3-relative threshold ***

@@

	\section{Summary}
Lattice matching based on unit cell parameters and 3-dimensional
transformation matrices is described, including the production
of the matrices for transforming unit cell base vectors from
a probe's to those of a reference cell.

@@

	The code for Lattice Matching (CmdLMP3) and related 
	software tools is available in github.com, in
	\url{https://github.com/duck10/LatticeRepLib.git}.
	The program CmdLMP3 uses the required files.

@@

	%\section{blah blah blah -- Supplementary Material}

\ack{{\bf Acknowledgements}}

Careful copy-editing and corrections by Frances C. Bernstein are 
gratefully acknowledged. M.I. Aroyo kindly provided test examples,  \citeasnoun{aroyo2025pc}\\.

\ack{{\bf Funding information}}

Funding for this research was provided in part by:  
US Department of Energy Offices of Biological and 
Environmental Research and of Basic Energy Sciences 
(contract No. DE-AC02-98CH10886; contract No. E-SC0012704); 
U.S. National Institutes of Health (grant No. P41RR012408; 
grant No. P41GM103473; grant No. P41GM111244; 
grant No. R01GM117126,
grant No. 1R21GM129570); Dectris, Ltd.

@@

A solution is proposed for the problem of best matching the 
lattices of unit cells that are described in different
presentations along with the 3D matrix
that defines the relationship between matched
lattices.
